---
title: "Class 09: Candy Mini Project"
author: "Barry (PID: 911)"
format: pdf
toc: true
---

## Background
Today we are taking a wee detour to analyze a fun data-set (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA)

## Data import
The data is all related to Halloween candy and is from the 538 website.

```{r}
candy <- read.csv("candy-data.txt", row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

> Q2. How many fruity candy types are in the dataset?

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
library(dplyr)

candy |>
  filter(row.names(candy)=="Air Heads") |>
  select(pricepercent, sugarpercent, winpercent)
  
```

> Q4. What is the winpercent value for “Kit Kat”?

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

There is a useful `skim()` function in the **skimr** package that can help give you a quick overview of a given dataset. Let’s install this package and try it on our candy data.

```{r}
skimr::skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

> Q7. What do you think a zero and one represent for the candy$chocolate column?

## Exploratory analysis

> Q8. Plot a histogram of `winpercent` values using both base R and ggplot2.

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=10, fill="steelblue")
```

> Q9. Is the distribution of winpercent values symmetrical?

No

> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```
```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?


```{r}
choc.ind   <- as.logical(candy$chocolate)
choc.candy <- candy[choc.ind, ]
choc.win   <- choc.candy$winpercent
mean(choc.win)
```

```{r}
fruit.ind <- as.logical(candy$fruity)
fruit.candy <- candy[fruit.ind,]
fruit.win <- fruit.candy$winpercent
mean(fruit.win)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```


## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
candy |>
  arrange(winpercent) |>
  select(winpercent) |>
  head(5)
```


```{r}
x <- c(5,10,1)
sort(x)
```
```{r}
x <- c(5,10,1)
order(x)
```
```{r}
x[ order(x) ]
```

```{r}
head( candy[ order(candy$winpercent), ] )
```


> Q14. What are the top 5 all time favorite candy types out of this set?

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```


Adding some color

```{r, fig.height=15, fig.width=10}
ggplot(candy) +
  aes(winpercent, 
      reorder(rownames(candy), winpercent),
      fill=fruity) +
  geom_col()
```

```{r, fig.height=15, fig.width=10}
mycols <- rep("black", nrow(candy))

## chocolate candy in chocolate color
mycols[ as.logical(candy$chocolate) ] <- "chocolate"
# candy bars in brown
mycols[ as.logical(candy$bar) ] <- "brown"
# fruity candy in pink
mycols[ as.logical(candy$fruity) ] <- "pink"


ggplot(candy) +
  aes(winpercent, 
      reorder(rownames(candy), winpercent)) +
  geom_col(fill=mycols)
```


## Taking a look at pricepercent

Make a plot of winpercent vs pricepercent

```{r}
ggplot(candy) +
  aes(winpercent,
      pricepercent,
      label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text()
```

We can fix the label text overplotting with an add-on package called **ggrepel** and it's `geom_text_repel()`

```{r}
# fruity candy in pink
mycols[ as.logical(candy$fruity) ] <- "red" #"pink"

```

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent,
      pricepercent,
      label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols, max.overlaps = 5)
```

## Exploring the correlation structure

We can calculate the pair-wise correlation of all our columns

```{r}
cij <- cor(candy)

library(corrplot)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?


## Principal Component Analysis

In this case we want to be sure to set `scale=TRUE` argument for `prcomp()` because we have one variable, `winpercent`, that is on a very different scale than all others and would otherwise dominate our PCA results.


```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

First major result figure is the "score plot" of PC1 vs PC2 - how the different candy are related to each other on our new PC axis:

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point(col=mycols) +
  geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)
```

The second major results figure from PCA is the so-called "loadings plot"

```{r}
ggplot(pca$rotation) +
  aes(PC1, 
      reorder( row.names(pca$rotation), PC1)) +
  geom_col()
```


