---
title: "Class 08: Breast Cancer Mini Project"
author: "Barry (PID: 911)"
format: pdf
toc: true
---

## Background

In today's class we will be employing all the R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy data.

## Data Import

The data is in CSV format:

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
```

wee peak at the data
```{r}
head(wisc.df, 3)
```
> Q1. How many observations are in this dataset?


> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum( wisc.df$diagnosis == "M" )
```
```{r}
table( wisc.df$diagnosis )

```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}

length(grep("_mean", colnames(wisc.df)) )
```
```{r}
colnames(wisc.df)
```

We need to remove the `diagnosis` column before we do any further analysis of this dataset - we don't want to pass this to PCA etc. We will save it as a separate wee vector that we can use later to compare our findings to those of experts.

```{r}
wisc.data <- wisc.df[ ,-1]
diagnosis <- wisc.df$diagnosis
```

## Principal Component Analysis (PCA)

The main function in base R is called `prcomp()` we will use the optional argument `scale=TRUE` here as the data columns/features/dimensions are on very different scales in the origional data set.

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
```

```{r}
attributes(wisc.pr)
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

```{r}
summary(wisc.pr)
```

### Scree plot

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?


```{r}
wisc.pr$rotation["concave.points_mean", "PC1"]
```

# 4. Hierarchical clustering

The goal of this section is to do hierarchical clustering of the original data to see if there is any obvious grouping into malignant and benign clusters.

In short, these results are not good!

First we will scale our `wisc.data` then calculate a distance matrix, then pass to `hclust()`:

```{r}
wisc.dist <- dist( scale(wisc.data) )
wisc.hclust <- hclust(wisc.dist)
plot(wisc.hclust)
```



```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=2)
table(wisc.hclust.clusters)
```
  
## Combining methods

The idea here is that I can take my new variables (i.e. the scores on the PCs `wisc.pr$x`) that are better descriptors of the data-set than the original features (i.e. the 30 columns in `wisc.data`) and use these as a basis for clustering.

```{r}
pc.dist <- dist(wisc.pr$x[ ,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```
```{r}
table(diagnosis)
```

I can now run `table()` with both my clustering `grps` and the expert `diagnosis`

```{r}
table(grps, diagnosis)
```
Our cluster "1" has 179 "M" diagnosis
or cluster "2" has 333 "B" diagnosis

179 TP
24  FP
333 TN
33  FN

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```
Specificity: TN/(TN+FP)
```{r}
333/(333+24)
```

## Prediction

We can use our PCA model for prediction of new un-seen cases:



