---
title: "Class 8: Breast Cancer Analysis Mini Project"
author: "Barry (PID: 911)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is to explore a complete analysis using the unsupervised learning techniques covered in our last class. 

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.

## Data import

Data was downloaded from the class website as a CSV file.

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.df)
```

## Data exploration

The first column `diagnosis` is the expert opinion on the sample (i.e. patient FNA).

```{r}
head(wisc.df$diagnosis)
```

Remove the diagnosis from data for subsequent analysis
```{r}
wisc.data <- wisc.df[,-1]
dim(wisc.data)
```

Store the diagnosis as a vector for use later when we compare our results to those from experts in the field.

```{r}
diagnosis <- factor(wisc.df$diagnosis)
```

> Q1. How many observations are in this dataset?

There are `r nrow(wisc.data)` observations/patients in the dataset


> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#colnames(wisc.data)
length( grep("_mean", colnames(wisc.data)) )
```

## Principal Component Analysis (PCA)

The `prcomp()` function to do PCA has a `scale=FALSE` default. In general we nearrly always want to set this to TRUE so our analysis is not dominated by columns/variables in our dataset that have high standard deviation and mean when compared to others just because the units of measurments are on different units/scales.


```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE )
summary(wisc.pr)
```

### PCA Score Plot
The main PC result figure is called a "score plot" or "PC plot" or "ordination plot"...

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

## PCA Scree-plot

A plot of how much variance each PC captures. We can get this from `wisc.pr$sdev` or from the output of `summary(wisc.pr)`

```{r}
var.tbl <- summary(wisc.pr)
head(var.tbl$importance)
```

```{r}
var <- c(1, var.tbl$importance[2,])
cum.var <- c(0, var.tbl$importance[3,])

plot(var * 100, typ="b", 
     ylab="Percent Variance Captured",
     xlab="PC Number")
```

```{r}

plot(cum.var * 100, typ="o", 
     ylab="Percent Variance Captured",
     xlab="PC Number",
     ylim=c(0,100))
points(var * 100, col="blue", typ="o")
```

### Communicating PCA results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean", "PC1"]
```

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

We need 5 PCs to capture more than 80% variance

```{r}
summary(wisc.pr)
```

## Hierarchical clustering

Just clustering the origional data is not very informative or helpful.

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist)
```

View the clustering dendrogram result

```{r}
plot(wisc.hclust)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table( wisc.hclust.clusters )
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```

## Combining methods (PCA and Clustering)

Clustering the origional data was not very productive. The PCA results looked promising. Here we combine these methods by clustering from our PCA results. In other words "clustering in PC space"...


```{r}
## Take the first 3 PCs
dist.pc <- dist( wisc.pr$x[,1:3] )
wisc.pr.hclust <- hclust(dist.pc, method="ward.D2")
```

View the tree...
```{r}
plot(wisc.pr.hclust)
abline(h=70, col="red")
```

To get our clustering membership vector (i.e. our main clustering result) we "cut" the tree at a desired height or to yield a desired number of "k" groups.

```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```
How does this clustering grps compare to the expert diagnosis

```{r}
table(grps, diagnosis)
```
Sensitivity: TP/(TP+FN)
Specificity: TN/(TN+FN)


## 7. Prediction

We can use our PCA model for prediction with new input patient samples.




