---
title: "Class 08: Breast Cancer Mini Project"
author: "Barry (PID: 911)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class.

Today we will analyze a biopsy data-set from fine needle aspiration (FNA) of breast mass. 


## Data Import

The data is made available as a CSV file for download. We can read this in using `read.csv()`:

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
```

```{r}
head( wisc.df, 3 )
```

Make sure we remove or exclude the `diagnosis` column from the data-set that we use for further analysis - this is the expert diagnosis as either M or B:

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- as.factor( wisc.df$diagnosis )
```

## Exploratory data analysis

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```
> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

```{r}
sum( wisc.df$diagnosis == "M" )
```

> Q3. How many variables/features in the data are suffixed with `_mean`?

We can use the `grep()` function to help us here:
```{r}
colnames(wisc.data)
```

```{r}
length( grep("_mean", colnames(wisc.data), value = T) )
```

## Principal Component Analysis (PCA)

We need to scale our data before PCA with the `scale=TRUE` argument to `prcomp()`.

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Let's see the "PC score plot"
```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```


```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```


## Hierarchical clustering

```{r}
wisc.hclust <- hclust( dist( scale(wisc.data) ) )
plot(wisc.hclust)
```

## Combining methods

```{r}
d <- dist(wisc.pr$x[,1:4])
wisc.pr.hclust <- hclust(d, method="ward.D2")
plot(wisc.pr.hclust)
abline(h=70, col="red")
```


```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```
How does this clustering `grps` correspond to the expert `diagnosis`?

```{r}
table(diagnosis)
```

```{r}
table(diagnosis, grps)
```














