---
title: "Class 8: Machine Learning Mini Project"
author: "Barry"
format: pdf
---

In today's mini-project we will explore a complete analysis using the unsupervised learning techniques covered in class (clustering and PCA for now).

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set FNA breast biopsy data. 

## Data import

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
head(wisc.df)
```

Remove the diagnosis column and keep it in a separate vector for later.

```{r}
diagnosis <- as.factor(wisc.df[,1])
wisc.data <- wisc.df[,-1]
head(wisc.data)
```

## Exploratory data analysis

The first step of any data analysis, unsupervised or supervised, is to familiarize yourself with the data.

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

First find the column names
```{r}
colnames(wisc.data)
```

Next I need to search within the column names for "_mean" pattern. The `grep()` function might help here. 

```{r}
inds <- grep( "_mean", colnames(wisc.data))
length(inds)
```

> Q. How many dimensions are in this dataset?

```{r}
ncol(wisc.data)
```

# Principal Component Analysis

First do we need to scale the data before PCA or not.

```{r}
round( apply(wisc.data, 2, sd), 3)
```

Looks like wee need to scale.

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp( wisc.data, scale=TRUE )
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27% 

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs capture 72%

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?


## PC plot

We need to make our plot of PC1 vs PC2 (a.k.a score plot, PC-plot, etc.). The main result of PCA...

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=diagnosis)

```

```{r}
library(ggplot2)

pc <- as.data.frame(wisc.pr$x)
pc$diagnosis <- diagnosis

ggplot(pc) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

## Variance explained

We can get this from the outupt of the `summary()` function.

```{r}
summary(wisc.pr)
```


Calculate the variance of each principal component by squaring the sdev component of wisc.pr (i.e. wisc.pr$sdev^2). Save the result as an object called pr.var.

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

Calculate the variance explained by each principal component by dividing by the total variance explained of all principal components. 

```{r}
pve <- pr.var / sum(pr.var)
head(pve)
```
```{r}
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     type = "o")
```

# Examine the PC loadings

How much do the original variables contribute to the new PCs that we have calculated? To get at this data we can look at the the `$rotation` component of the returned PCA object.

```{r}
head(wisc.pr$rotation[,1:3])
```

Focus in on PC1

```{r}
head(wisc.pr$rotation[,1])
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean" ,1]
```

There is a complicated mix of variables that go together to make up PC1 - i.e. there are many of the original variables that together contribute highly to PC1.

```{r}
loadings <- as.data.frame(wisc.pr$rotation)

ggplot(loadings) +
  aes(PC1, rownames(loadings)) +
  geom_col()
```

# 3. Hierarchical clustering

The goal of this section is to do hierarchical clustering of the original data.

First we will scale the data, then distance matrix, then hclust 

```{r}
wisc.hclust <-  hclust( dist( scale(wisc.data) ) )
```

```{r}
plot(wisc.hclust)
```

Cut this tree to yield cluster membership vector with `cutree()` function.

```{r}
grps <- cutree(wisc.hclust, h=19)
table(grps)
```

```{r}
table(grps, diagnosis)
```

# Combine methods: PCA and HCLUST

My PCA results were intresting as they showed a seperation of M and B samples along PC1.

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=diagnosis)
```


I want to cluster my PCA results - that is use `wisc.pr$x` as input to `hclust()`.


Try clustering in 3 PCs, that is PC1, PC2 and PC3 as input
```{r}
d <- dist(wisc.pr$x[,1:3])

wisc.pr.hclust <- hclust(d, method="ward.D2")
```

An my tree result figure
```{r}
plot(wisc.pr.hclust)
```

Let's cut this tree into two groups/clusters

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=grps)
```

How well do the two clusters separate the M and B diagnosis?
```{r}
table(grps, diagnosis)

```

```{r}
(179+333)/nrow(wisc.data)
```



