---
title: "Class 6: R functions"
author: "Barry (PID: 911)"
format: gfm
---

All functions in R have at least 3 things:

- A **name**, we pick this and use it to call the function.
- Input **aruguments**, there can be multiple comma seperated inputs to the function.
- The **body**, lines of R code that do the work of the function.


Our first wee function:

```{r}
add <- function(x, y=1) {
  x + y
}
```

Let's test our function

```{r}
add(c(1,2,3), y=10 )
```

```{r}
add(10)
```


```{r}
add(10, 100)
```

## A second function

Let's try something more intresting. Make a sequence generation tool.

The `sample()` function could be useful here.

```{r}
sample(1:10, size = 3)
```
change this to work with the nucleotides A C G and T and return 3 of them

```{r}
n <- c("A","C","G","T")
sample(n, size=15, replace = TRUE)
```
Turn this snipet into a function that returns a user specifed length dna sequence. Let's call it `generate_dna()`...

```{r}
generate_dna <- function(len=10, fasta=FALSE) {
  n <- c("A","C","G","T")
  v <- sample(n, size=len, replace = TRUE)

  # Make a single element vector
  s <- paste(v, collapse="")

  cat("Well done you!\n")

  if(fasta) {
    return( s )
  } else {
    return( v )
  }
}
```

```{r}
generate_dna(5)
```
```{r}
s <- generate_dna(15)
s
```
I want the option to return a singe elemet character vector with my sequence all together like this: "GGAGTAC"


```{r}
generate_dna(10, fasta=TRUE)
```
```{r}
generate_dna(10, fasta=FALSE)

```

## A more advanced example

Make a thrid function that generates protein sequence of a user specifed lenght and format.


```{r}
generate_protein <- function(size=15, fasta=TRUE) {
  aa <- c("A", "R", "N", "D", "C", "Q", "E", "G", 
          "H", "I", "L", "K", "M", "F", "P", "S", 
          "T", "W", "Y", "V")
  
  seq <- sample(aa, size = size, replace = TRUE)

  if (fasta) {
    return(paste(seq, collapse = ""))
  } else {
    return(seq)
  }
}
```
Try this out...
```{r}
generate_protein(10)
```
> Q. Generate random protein sequnces between lengths 5 and 12 amino-acids.

```{r}
generate_protein(5)
generate_protein(6)

```
One apprach is to do this by brute force calling our function for each length 5 to 12.

Another apprach is to write a `for()` loop to itterate over the input valued 5 to 12

A very useful third R specific apprach is to use the `sapply()` function.

```{r}
seq_lengths <- 6:12
for (i in seq_lengths) {
  cat(">",i, "\n", sep="")
  cat( generate_protein(i) )
  cat("\n")
}
```

```{r}
sapply(5:12, generate_protein)
```

> **Key-Point**: Writing functions in R is doable but not the eaisest thing. Starting with a working snippet of code and then using LLM tools to improve and generalize your funtion code is a productive approach.



