---
title: "Class 6: R functions"
author: "Barry (PID: 911)"
format: pdf
toc: true
---

## Background

All functions in R have at least 3 things:

-   a *name* (we pick that and use it to call the function)
-   input *arguments* (one or more comma separated inputs that go inside the brackets when we call the function),
-   the *body* (the lines of R code that do the work of the function).

## A first function

Here we will create a function to add some numbers. Let's call it `add()`.

Input arguments can be either **"required"** or **"optional"**. The later have fall-back default values that will be used if the user does not specify them.

> **Q1a**. Your first version of the function should add two input numbers together. For example, add(4, 7) should return 11. \[1 pt\]

```{r}
add <- function(x, y=0) {
  x + y
}
```

Can we use our new function:

```{r}
add(10, 100)
```

```{r}
add(10)
```

> **Q1b**: For you second version, adapt your first function so it can take a single input vector or two inputs as before. For example, add(4, 7) and add( c(4,7,10) ). \[1 pt\]

```{r}
add <- function(x, y=0) {
  sum(x,y)
}
```

```{r}
add(4, 7)
add( c(4,7,10))
```

> **Q1c**. To do on your own... create a third version of your function that can add any number of inputs that the user provides. For example, add(1, 2, 3, -4)

```{r}
add <- function(x, y=0, z=0) {
  sum(x,y,z)
}
```

```{r}
add(x=1, y=2, z=13)
```

```{r}
ans <- add(1, 2, 3)
ans
```

We can explicitly set a **return** value output from a function (rather than the default last line) by using the `return()` function call.

```{r}
add <- function(x, y=0, z=0) {
  return(sum(x,y,z))
  cat("Is it break time yet?\n")
}

add(10,100)
```

## A generate_dna() function

A useful function here is the "base R" `sample()` function:

```{r}
sample(1:5, size=3)
```

```{r}
sample(1:5, size=60, replace = TRUE)
```

We can use this to make a random nucleotide sequence if we draw from "A", "C", "G" and "T"...

```{r}
sample(x=c("A","C","G","T"), size=10, replace = TRUE)
```

> **Q2a**: Write a function `generate_dna()` that returns a random DNA sequence of a length specified by the user. Your first version should return a multi-element vector of single character nucleotides. For example generate_dna(6) might return "A", "T", "T", "G", "A", "C". \[1 pt\]

```{r}
generate_dna <- function(len=10) {
  sample(x=c("A","C","G","T"), size=len, replace = TRUE)
}
```

```{r}
generate_dna()
generate_dna(len=100)
```

> **Q2b**: Your second version should *optionally* be able to return either a multi-element vector of single character nucleotides (as before) or a **single character string** (not a vector of single letters but a singe vector of multiple letters). For example "AAGGCTG". \[1 pt\]

```{r}
generate_dna <- function(len=10, single.element=TRUE) {

  ans <- sample(x=c("A","C","G","T"), size=len, replace = TRUE)

  #cat("Hello....")
  
  if(single.element) {
    # cat("is it me you are looking for...")
    ans <- paste( ans, collapse = "" )
  }
  return(ans)
}
```

Functions that could be useful here are `paste()`, `if()`, `cat()` and `return()`

```{r}
generate_dna()
generate_dna(single.element = FALSE)
```

> **Q2c**. Finally, create a final version of your function that prints out a FASTA format sequence with an id line indicating the sequence length.

```         
>len9
CGAAGGCTG
```

```{r}
cat("hello \n there")
```

```{r}
generate_dna <- function(len=10, single.element=TRUE) {

  ans <- sample(x=c("A","C","G","T"), size=len, replace = TRUE)

  if(single.element) {
     ans <- paste( ans, collapse = "" )
  }
  
  ## Format as FASTA with an ID line
  cat( paste(">len", len, "\n", sep="") )
  cat(ans)
  cat("\n")
  ## 
  
  return(ans)
}
```

```{r}
x <- generate_dna(20)
```

## Write a `generate_protein()` function

> **Q3**. Write a function `generate_protein()` that returns a random peptide/protein sequence of a length specified by the user. For example `generate_protein(6)` might return `"WQRTAG"`.

```{r}
generate_protein <- function(len=9) {
  aa <- c("A", "R", "N", "D", "C", "E", "Q", 
          "G", "H", "I", "L", "K", "M", "F", 
          "P", "S","T", "W", "Y", "V")
  
  ans <- sample(x=aa, size=len, replace = TRUE)
  paste(ans, collapse = "")
}
```

```{r}
generate_protein(5)
```

## Generate random protein sequences of length 6 to 13

> **Q4** Adapt and use your `generate_protein()` function to generate a series of random protein sequences ranging from 6 to 13 amino acids in length (one sequence per length). Take advantage of the base R function `for()` or `sapply()` so that you do not have to call generate_protein() eight times by hand.

```{r}
for(l in 6:13) {
  cat(">", l, "\n", sep="")
  cat( generate_protein(l), "\n" )
}
```

```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
```

## Are Our peptides “unique in nature”?

> **Q5**: Take your FASTA-formatted peptides from Q4 and run them as a single BLASTp search against the Non-redundant protein sequences (nr) database at https://blast.ncbi.nlm.nih.gov/. For this question do not restrict the organism (leave the Organism field blank so that the full nr database is searched).

| length | Ide | Cov | Unique |
|------|------|------|------|
|   6  |  100 | 100  |  N   |
|   7  |  100 | 100  |  N   |
|   8  |  100 |  88  |  Y   |
|   9  |  100 |  89  |  Y   |
|   10 |   90 | 100  |  Y   |
|   11 |   90 |  91  |  Y   |
|   12 |   77 | 100  |  Y   |
|   13 |  100 |  69  |  Y   |


## Connecting our findings to immunology

> **Q6**: In 3–6 sentences total and using your Q5 data and the reasoning from Q5b, what do you think this minimum length is and why might it be a bad design choice for the immune system to present very short peptides? [3 pt]


