---
title: "Class 7: Machine Learning 1"
author: "Barry (PID: 911)"
format: html
toc: true
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionallity reduction**.


## K-means clustering

The main function in "base" R for K-means clustering is called `kmeans()`

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To do this we can use the `rnorm()` function:


```{r}
hist( rnorm(10000, mean=3) )
```

```{r}
x <- c( rnorm(30, -3), rnorm(30, +3) )

z <- cbind(x=x, y=rev(x))
plot(z)

```

Now we can run `kmeans()` on this input `z` and see what the results look like.

```{r}
km <- kmeans(z, centers = 2)
km
```

```{r}
attributes(km)
```

> Q. How many points are in each cluster?

```{r}
km$size
```


> Q. What "component of your result object details cluster assignment/membership? 

```{r}
km$cluster
```

> Q. What "component of your result object details cluster center?

```{r}
km$centers
```


> Q. Plot `z` colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
plot(z, col=c("red","blue") )
```


```{r}
plot(z, col=km$cluster)
points(km$centers, col="blue", pch=15)
```

> Q. Run a K-means clustering and plot the results asking for 4 clusters (K=4)?

```{r}
km4 <- kmeans(z, centers = 4)
plot(z, col=km4$cluster)
points(km4$centers, col="black", pch=15)
```

> **N.B.** You need to tell K-means the number of clusters (i.e. set `centers=2`)!!

One approach is to try different values for `centers` and then pick the best...

```{r}
ans <- NULL
for(i in 1:10) {
  km <- kmeans(z, centers=i)
  ans <- c(ans, km$tot.withinss)
}

plot(ans, typ="o", 
     xlab="Number of clusters",
     ylab="Total Sum of Squars Distance")
```

## Hierarchical Clustering

The main function in "base" R for Hierarchical Clustering is called `hclust()`

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()`


```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

There is a bespoke `plot()` method for `hclust()` result objects.

```{r}
plot(hc)
abline(h=8, col="red")
```

Once we have our `hclust` object (our "tree" of "cluster dendrogram") we can *"cut"* the tree to reval the clustering pattern.

```{r}
cutree(hc, h=8)
```


```{r}
cutree(hc, k=4)
```

> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership)

```{r}
grps <- cutree(hc, k=2)
plot(z, col=grps)
```


## Principal Component Analysis (PCA)

PCA is a dimensionallity reduction method that is popular for revealing patterns in complex datasets.

### Analysis of UK food data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regions being distinct from others. 

## Data Import

The data is made availabe in CSV format so we can use the `read.csv()` function for import to R:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

```{r}
row.names(x) <- x[,1]
x <- x[,-1]
x
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?


```{r}
x <- read.csv(url, row.names=1)
```


## Tidy the data

Fix anything that went wrong with data import.

```{r}

```


> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?


## Exporatory analysis

Make some plots to help make sense of obvious trends...


> Q5. We can use the `pairs()` function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```

> **Key-point**: Even relatively small datasets can prove challenging to interpret. 

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`. This function wants the "observations" to be rows and the "variables" to be columns. 

So here we need to take the transpose of our `x` input object

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

The returned `pca` object has components that we can use to make our main result figures:

```{r}
attributes(pca)
```
The main result figure from this analysis is called a **"PC score plot"** (a.k.a. an "ordination plot", "PC plot" or simply "PC1 vs PC2 plot").

This plot shows how samples (in this case countries) relate to each other along our new PC axis. 

This is our new "reduced-dimensional space". In this case 2 dimensions, PC1 and PC2, that capture most of the variance in the original 17 dimensional data-set.

```{r}
library(ggplot2)

ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point() 
```

```{r}
mycols <- c("orange", "red", "blue", "darkgreen")

ggplot(pca$x) +
  aes(PC1, PC2) +
  geom_point(col=mycols)
```

```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point(col=mycols) +
  geom_text(size=3, vjust=2, col=mycols)
```


```{r}
ggplot(pca$x) +
  aes(PC1, PC2, label=row.names(pca$x)) +
  geom_point() +
  geom_text(size = 2,vjust = 2)

```

```{r}
ggplot(pca$rotation) +
  aes(PC1, 
      reorder( row.names(pca$rotation), PC1) ) +
  geom_col()
```




