---
title: "Class 7: Machine Learning I"
author: "Barry"
format: html
---

In this class we will explore clustering and dimensionality reduction methods.

## K-means

Make up some input data where we know what the answer should be.

```{r}
tmp <- c( rnorm(30, -3), rnorm(30, +3) )
x <- cbind(x=tmp, y=rev(tmp))
head(x)
```

Quick plot of x to see the two gropus at -3,+3 and +3,-3
```{r}
plot(x)
```

Use the `kmeans()` function setting k to 2 and nstart=20 

```{r}
km <- kmeans(x, centers = 2, nstart=20)
km
```



> Q. How many points are in each cluster?

```{r}
km$size
```


> Q. What ‘component’ of your result object details 
      - cluster assignment/membership? 
      - cluster center?

```{r}
km$cluster
km$centers
```

> Q. Plot x colored by the kmeans cluster assignment and 
      add cluster centers as blue points

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15, cex=2)
```

Using ggplot requires data frames as input:

```{r}
x.df <- as.data.frame(x)
x.df$cluster <- as.factor(km$cluster)

library(ggplot2)

ggplot(x.df) +
  aes(x, y, col=cluster) +
  geom_point()
```

Play with kmeans and ask for different number of clusters
```{r}
km <- kmeans(x, centers = 4, nstart=20)
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=16, cex=2)

```

# Hierarchical Clustering

This is another very useful and widely employed clustering method which has the advantage over k-means in that it can help reveal the something of the true grouping in your data. 

The `hclust()` function wants a distance matrix as input. We can get this from the `dist()` function.

```{r}
d <- dist(x)
hc <- hclust(d)
hc
```

There is a plot method for hclust results:

```{r}
plot(hc)
abline(h=10, col="red")
```

To get my cluster membership vector I need to "cut" my tree to yield sub-trees or branches with all the members of a given cluster residing on the same cut branch. The function to do this is called `cutree()`

```{r}
grps <- cutree(hc , h=10)
grps
```

```{r}
plot(x, col=grps)
```
```{r}
plot(hc)
```

It is often helpful to use the `k=` argument to cutree rather than the `h=` height of cutting with `cutree()`. This will cut the tree to yield the number of clusters you want.


```{r}
cutree(hc, k=4)
```

# Principal Component Analysis (PCA)

The base R function for PCA is called `prcomp()` 
Let's play with some 17D data (a very small dataset) and see how PCA can help.

## PCA of UK food data

Import the data

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
head(x)
```

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

A "PCA plot" (a.k.a "Score plot", PC1vsPC2 plot, etc.)

```{r}
pca$x
```

```{r}
plot(pca$x[,1], pca$x[,2], 
     col=c("orange", "red", "blue", "darkgreen"), pch=15)
```

```{r}
pc <- as.data.frame(pca$x)
pc$contry <- row.names(pca$x)

library(ggplot2)

ggplot(pc) +
  aes(PC1, PC2, col=contry, label=contry) +
  geom_point() +
  geom_text()
```



# Tidymodels approach

```{r}
library(tidymodels)
```

```{r}
summary(km)
```

The `tidy()` function summarizes on a per-cluster level:

```{r}
tidy(km)
```

And as it always does, the glance() function extracts a single-row summary:

```{r}
glance(km)
```


```{r}
# The answer we want is in km$tot.withinss

nrun <- 1:9
tot <- NULL

for(i in nrun) {
  tot <- c(tot, kmeans(x, centers = i)$tot.withinss)
}

```

And a scree-plot 
```{r}
plot(tot, typ="o")
```

```{r}
head( tidy(d) )
```

```{r}
glance(pca)
```


```{r}
loadings <- as.data.frame(pca$rotation)

cols <- rep("gray",17)
cols[loadings$PC1 > 0.05] <- "darkgreen"
cols[loadings$PC1 < -0.25] <- "orange"
  
  
ggplot(loadings) +
  aes(PC1, rownames(loadings)) +
  geom_col(fill=cols) +
  theme_bw()
```



# Exploratory clustering

Let’s say we want to explore the effect of different choices of k, from 1 to 9, on this clustering. First cluster the data 9 times, each using a different value of k, then create columns containing the tidied, glanced and augmented data:


# TODO: Figure out map function

```{r}
df <- data.frame(x=tmp, y=rev(tmp))

kdf <- kmeans(df, centers = 2)
kdf
```

```{r}
ggplot(df) +
  aes(x, y) +
  geom_point(col=kdf$cluster)
```

```{r}
hdf <- hclust(dist(df))
plot(hdf)
```



```{r}
points <- x

kclusts <- 
  tibble(k = 1:9) %>%
  mutate(
    kclust = map(k, ~kmeans(points, .x)),
    tidied = map(kclust, tidy),
    glanced = map(kclust, glance),
    augmented = map(kclust, augment, points)
  )

kclusts
```

We can turn these into three separate data sets each representing a different type of data: using tidy(), using augment(), and using glance(). Each of these goes into a separate data set as they represent different types of data.


```{r}
clusters <- 
  kclusts %>%
  unnest(cols = c(tidied))

assignments <- 
  kclusts %>% 
  unnest(cols = c(augmented))

clusterings <- 
  kclusts %>%
  unnest(cols = c(glanced))
```

Now we can plot the original points using the data from augment(), with each point colored according to the predicted cluster.

```{r}
p1 <- 
  ggplot(assignments, aes(x, y)) +
  geom_point(aes(color = .cluster), alpha = 0.8) + 
  facet_wrap(~ k)
p1
```

```{r}
p2 <- p1 + geom_point(data = clusters, size = 10, shape = "x")
p2
```

The data from glance() fills a different but equally important purpose; it lets us view trends of some summary statistics across values of k. Of particular interest is the total within sum of squares, saved in the tot.withinss column.

```{r}
ggplot(clusterings, aes(k, tot.withinss)) +
  geom_line() +
  geom_point()
```

This represents the variance within the clusters. It decreases as k increases, but notice a bend (or “elbow”) around k = 2. This bend indicates that additional clusters beyond the third have little value. (See here for a more mathematically rigorous interpretation and implementation of this method). Thus, all three methods of tidying data provided by broom are useful for summarizing clustering output.



