---
title: "Bio3D intro"
author: "barry"
format: html
---


# Getting started with Bio3D

Bio3D is an R package for structural bioinformatics available on CRAN

We will also use a couple of other R packages from GitHub and bioconductor in this lab.


## Setup some needed packages

```r
# Pak can install from multiple sources
install.packages("pak")

pak::pkg_install( c("bioboot/bio3dview", # GitHub
                    "NGLVieweR",         # CRAN
                    "bioc::msa") )       # Bioconductor
```

## Reading PDB file data into R


```{r}
library(bio3d)

pdb <- read.pdb("1hsg")
```

The returned `pdb` object is a list with all the key structural information now accessible in R:

```{r}
head(pdb$atom)
```

There are many functions that work on these `pdb` objects, e.g.

```{r}
head( pdbseq(pdb) )
```

Let's try a new (not on CRAN yet) package for quick interactive structure viewing:

```{r}
library(bio3dview)

view.pdb(pdb)
```
```{r}
library(NGLVieweR)

view.pdb(pdb, 
         colorScheme = "sse", 
         backgroundColor = "black") |>
  setRock()
```
 > Q. Can you make a highly cusom interactive view of this HIV-Pr structure highligthing the active site ASP residues (`resno=25`), the drug/ligand, the two chains in your choice of colors, along with a custom background color?

```{r}
#| eval: !expr knitr::is_html_output()


active.site <- atom.select(pdb, resno=25)

view.pdb(pdb, 
         highlight = active.site, 
         highlight.style = "spacefill",
         cols=c("orange","teal"), 
         backgroundColor = "pink")
```

## Prediction of flexiblity

In addition to viewing and storing structure data we can do various bioinformatics calculations such as a NMA to predict flexiblity and inernal dynamics:

```{r}
adk <- read.pdb("6s36")
adk
```

```{r}
m <- nma(adk)
plot(m)
```

```{r}
view.nma(m)
```

```{r}
mktrj(m, file="nma.pdb")
```

